---
title: "Capital Asset Pricing model"
author: "EL KAMLI Adam"
output:
  pdf_document:
    toc: yes
  word_document:
    toc: yes
  html_document:
    toc: yes
    toc_float: yes
    highlight: pygments
    theme: united
    numbere_sections: yes
---

```{r, warning=FALSE, message=FALSE}
library(tidyverse) #Data analytics
library(xts) #Xts objects (xtensible time series)
library(quantmod)#Quantitative function
library(readxl)#Read excel files
library(zoo) #Zoo objects(time series)

```

# Introduction

Factor models are used in many financial applications, such as identifying the determinants of a security's return as well as in cost of capital calculations. Factors help explain the variation in a security's return. The simplest factor model is one based on a single factor.

The most popular of these models is the Capital Asset Pricing Model (CAPM). In the CAPM, only the sensitivity of the security's return to the market portfolio's return matters.

In reality the CAPM does not preform well in empirical testing, and one of the reasons is the difficulty to use a \*real market portfolio\*.

Thus in empirical studies, we use a more simple one single factor know in some literature as the \*market model\* that resembles the CAPM model but with no imposition on what kind of market proxy to use, and thus in such model we can use in broad-based market index.

# C.A.P.M

The value of a company can be seen as the seen as the present value of dividend payments that the shareholders except to receive from holding the stock.

Since the dividend are expected to arrive at some future dates, we need to calculate the discount rate at which we will discount those streams.

That risk-adjusted rate is given by the CAPM model as :

$$
r_i = r_f + \beta_i(r_m - r_f)
$$

where :

-   $r_i$ the risk adjusted rate.

-   $r_f$ the risk free rate.

-   $\beta_i$ the beta which measures the sensitivty of the asset $i$ to the overall market.

-   $r_m$ return on the market proxy.

## CAPM regression

In empirical studies however this equation tales another form :

$$
r_i - r_f = \alpha+\beta_i(r_f-r_m)
$$

Three parameters interfere in this regression :

1.  The length of the estimation period
2.  The frequency of the returns data
3.  The risk-free rate used

In this project we are gonna take 36 months returns rate and the Morocco 3-Month Bond Yield as the risk free rate of return.

```{r}
port <- read.csv("Portfolio(monthly).csv")
head(port, n=2)
tail(port, n=2)

```

```{r}
#we converte date to it's original format
port$date <- as.yearmon(port$date, "%b %Y")
port <- as.data.frame(port)
head(port)

```

Next we import the MASI (Moroccan All Shares Index) which is the moroccan market proxy).

```{r}
MASI <- read.csv("Moroccan All Shares - Données Historiques.csv")

masi <- MASI[,c(1,2,3,4,5)]
masi$Date <- as.Date(masi$Date, format = "%d/%m/%Y")
head(masi)


```

```{r}
#As we can see the numerical values in this data have a peculiar form 
clean_transform <- function(x){
    x <- gsub(",", '.', x)
    x <- gsub("\\.", "", x)
    return(as.numeric(x))
}
clean_and_convert <- function(x) {
  x <- gsub("[^0-9.-]", "", x)  
  as.numeric(x)
}

#We apply this function to all the columns
masi[,-1] <- lapply(masi[,-1], clean_transform)
names(masi) <- paste(c("date", "mrkt.Close", "mrkt.Open", "mrkt.High", "mrkt.Low"))
masi <- masi[,c(1,3,4,5,2)]
head(masi)

```

```{r}
#We convert the data to xts
masi <- as.xts(masi[,-1], order.by = masi$date)
masi.monthly <- to.monthly(masi)
masi.monthly <- masi.monthly[,4]
head(masi.monthly)
masi.ret <- Delt(masi.monthly$masi.Close)
names(masi.ret) <- paste(c("mrkt.ret"))
masi.ret <- masi.ret[-1,]
head(masi.ret)

```

```{r}
#For compatibility with the rest of the data we are gonna convert masi.ret to a dataframe
market.df <- as.data.frame(masi.ret)
head(market.df)
```

Now that we have our market proxy all set and done, we are gonna import our risk free rates

```{r}
rf <- read.csv("Morocco 3-Month Bond Yield Historical Data.csv")
head(rf)

```

```{r}
rf <- rf[, c(1,2)]
rf$Date <- as.Date(rf$Date,format = "%m/%d/%Y")
rf$Price <- clean_and_convert(rf$Price)
head(rf)
```

```{r}

rf <- as.xts(rf, order.by = rf$Date)
rf <- rf[,-1]
rf <- xts(apply(rf, 2, as.numeric), order.by = index(rf))
rf.monthly <- to.monthly(rf)

```

Now we should convert the annualized yield of each month, to a monthly yield using the following formula :

$$
rf_{monthly} = \left(1+\frac{rf_{annualized}}{100}\right)^{1/12}-1
$$

```{r}
rf.monthly<-(1+rf.monthly[,1]/100)^(1/12)-1
rf.monthly <- rf.monthly[-1,]
head(rf.monthly)
rf
```

Now we combine all the above datasets to one dataframe

```{r}
data <- cbind(market.df,data.frame(rf.monthly), port$port.ret)
names(data) <- paste(c("mkt.ret", "rf", "port.ret"))
head(data)
```

No we calculate the following :

-   $r_i - r_f$ : excess firm return.
-   $r_m - r_f$ : excess market return.

```{r}
data$exret <- data$port.ret-data$rf
data$exmkt <- data$mkt.ret-data$rf
head(data)
```

Now we run our regreesion

```{r}
capm <- lm(exret~exmkt, data = data)
summary(capm)
```

We find that :

$$
r_i-r_f = 0.016833+0.631525(r_m-r_f)
$$

```{r, message=F}
ggplot(data, aes(x = exret, y = exmkt))+
  geom_point(color = 'blue', size = 2)+
    geom_smooth(method = "lm", color = 'red')+
    labs(title = "Scatter plot of the exret Vs. exmkt market return with regression line",
         x = "excedent return",
         y = "edxcedent market return")+
  theme_minimal()
```

We can see from this graph why the CAPM is not the best fitted for empirical studies, however it can gives us some good insights.\

## Coefficient interpretation

### Alpha :

When putting our money into a fund we want to know if the fund manager has a plus value or does he just invest in the index .

To do so we look at his performance compared to a benchmark, by looking at his *Alpha* since the performance of the market is measured in the *Beta* then any excess value can be attribute to the manager.

If the *Alpha* is statistically significant then it can interpreted as the positive or negative value of the manager, in our example, the manager has recorder an incremental return of **16%** and it's statistically significant to *5%* level which is the confidence level agreed upon.

### Beta :

The *Beta* on the other hand measures the sensitivity of the portfolio to the overall market movement.

In financial term we say that *Beta* measures the portfolio's **systematic risk** also called **market risk**, it's the portion of the risk that cannot be diversified.

In our example we have $\beta = 0.631525$ which means that if the market goes up by *1%* our portfolio will go up by *0.63%* and if it goes down by *1%* then the portfolio will go down by *0.63%*.

# Market model

Unlike the CAPM that requires *true* market portfolio, the market model uses the market proxy such as :

$$
r_i = \alpha +\beta r_m
$$

```{r}
market_model <- lm(port.ret~mkt.ret, data)
summary(market_model)

```

we can observe that the *beta* here is very close to that of the CAPM, and the interpretation are the same except that this time we compare it with the movement of the MASI directly.

```{r}
ggplot(data, aes( x = port.ret, y = mkt.ret))+
  geom_point(color = 'blue', size = 2)+
  geom_smooth(method = "lm", color = 'red')+
  labs ( title = "Scatter plot of port.ret Vs mkt.ret with a regression line fitted", 
         x = "Portfolio returns",
         y = "Market proxy returns")+
  theme_minimal()
```

We can see by comparing the two regression graphs that the slopes are indeed very similar.

# Rolling windows

As mentioned earlier the results of the regression on depends on the time window used, we give here an example of how the window frame can affect the values of *Alpha* and *Beta*.

```{r}
# We import the historical stock prices of LABEL-VIE
label <- read_excel('Label_vie.xlsx')
label$Séance <- as.Date(label$Séance, format = "%d/%m/%Y") #Format the date
label <- label[, c(1, 4, 5, 6, 7, 9, 12)] #Select the colomns that interest us
label[,-1] <- lapply(label[,-1], clean_and_convert) #We clean the data
label <- as.xts(label[, -1], order.by = label$Séance)#Convert to an xts object
colnames(label) <- c("LABEL.Open", "LABEL.Close", "LABEL.High", "LABEL.Low", "LABEL.Volume", "LABEL.Adjusted") #Standardise the names 
label <- label[, c(1, 3, 4, 2, 5, 6)]
head(label)

```

```{r}
# Now we calculate the market and LABEL vie logaithmic returns
rets <- diff(log(label$LABEL.Adjusted))
rets$masi <- diff(log(masi$mrkt.Close))
names(rets) <- paste(c("LABEL.ret", "MARKET.ret"))
rets <- na.omit(rets)
head(rets)

```

Now we run a regression over a financial year (252 days)

```{r}
reg <- function(X){
  roll.reg = lm(LABEL.ret ~MARKET.ret, data = as.data.frame(X))
  return(roll.reg$coef)
}
coeff <- rollapply(rets,
                   width = 252,
                   FUN = reg,
                   by.column = F)
coeff <- na.omit(coeff)
names(coeff) <- paste(c("Alpha", "Beta"))
head(coeff)
```

Now we plot our data to have a better idea on the evolution of our two coefficients.

```{r}

plot(x = index(coeff),
       y = coeff$Alpha,
       xlab = 'Date',
     ylab = 'Alpha',
     type = 'l',
     col = 'red',
     main = 'Label Vie. Alpha using 252-day rolling window'
     )
plot(x = index(coeff),
     y = coeff$Beta,
     xlab = "Date", 
     ylab = 'Beta',
     type = 'l',
     col = 'blue',
     main = 'Label Vie. Beta using 252-day rolling window')

```

As we can notice the Label Vie *Alpha* seems to fluctuate around very small values while the *Beta* takes on much bigger variations.
